Effects of pulsed hollow electron lens operation on the beam core in HL-LHC: \\First experimental studies and simulations.

### 2018-01-30 resex.tex
version 2018-01-29
first revised final version

\section{Summary\label{sec:sum}}
This paper intends to give an overview of the effects of the HEL on the proton beam core and the experiments conducted in this context at the LHC in 2016~\cite{resexmd2016} and 2017~\cite{resexmd2017}. For a perfect HEL without any imperfections no effect on the core is to be expected as the field at the center vanishes by construction. Only in the case of imperfections in the guidance and profile of the electron beam, a residual kick on the proton beam core can occur. In this case, the contribution from the electron lens bends with approximately $0.5$~nrad to first order is negligible to the kick expected in the central region of the HEL of approximatley $20$~nrad to first order caused by profile imperfections in the electron beam assuming HL-LHC parameters (Tables~\ref{tab:hllhc_param}--\ref{tab:hel_param}).

In DC operation these kicks, even if nonlinear in reality, are negligible compared to the nonlinearities present in the LHC and HL-LHC and are thus of no concern. The picture changes if the HEL is operated in pulsed mode, which introduces noise on the halo particles (intended) as well as on the beam core (unintended) in the presence of imperfections of the HEL entailing much more stringent tolerances. To study the effect on the beam core in case of the much more dangerous pulsed operation, beam experiments were conducted in 2016 and 2017 at injection. The experimental results in general suggest that for the kick amplitudes as expected from the HEL, considerable effects are to be expected for a random excitation and certain resonant excitations, explicitly $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing were tried during the experiments. In addition, also one resonant excitation pattern, the $8^{\mathrm{th}}$~turn pulsing, was tested featuring a small effect on the beam core. Assuming that this pattern has a large effect on the beam halo, it would be an optimal candidate for a feasible HEL excitation pattern.

The resonant excitation in general features a rich nonlinear dynamics and diversity in beam distribution and diffusion changes as can be seen from the very diverse effects of the different $k^{\mathrm{th}}$~turn pulsing patterns. The origin of this diversity is the fact that only certain frequencies and thus resonances are excited. Typical for the resonant excitation is also the fast adjustment of the beam distribution to a new equilibrium state which could be for the first time measured in the experiments and predicted in the simulations presented in this paper. The random excitation on the other hand excites all beam frequencies equally, leading to the highest increase in halo removal rates for the HEL \cite{hel_halo_hllhc_fitterer} but also the strongest effects on the core, mainly in terms of strong emittance growth as predicted in simulations and also a multitude of earlier publications on the subject of random noise.

>>>>

\section{Summary\label{sec:sum}}
This paper intends to give an overview of the effects of the HEL on the proton beam core and the experiments conducted in this context at the LHC in 2016~\cite{resexmd2016} and 2017~\cite{resexmd2017}. For a perfect HEL without any imperfections no effect on the core is to be expected as the field at the center vanishes by construction. Only in the case of imperfections in the guidance and profile of the electron beam, a residual kick on the proton beam core can occur. In this case and assuming HL-LHC parameters (Tables~\ref{tab:hllhc_param}--\ref{tab:hel_param}), the contribution from the electron lens bends with approximately $0.5$~nrad to first order is negligible with respect to the kick expected from the central region of the HEL of approximatley $20$~nrad to first order caused by profile imperfections in the electron beam.

In DC operation these kicks, even if nonlinear in reality, are negligible compared to the nonlinearities present in the LHC and HL-LHC and are thus of no concern. The picture changes if the HEL is operated in pulsed mode, in which case noise is introduced on the halo particles (intended) as well as on the beam core (unintended) in the presence of imperfections of the HEL entailing much more stringent tolerances.

To study the effect on the beam core in case of the much more dangerous pulsed operation, beam experiments were conducted in 2016 and 2017 at injection. The experimental results in general suggest that for the kick amplitudes as expected from the HEL, considerable effects are to be expected for a random excitation and certain resonant excitations, explicitly $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing during the conducted experiments. In addition, also one resonant excitation pattern, the $8^{\mathrm{th}}$~turn pulsing, was tested featuring a small effect on the beam core. Assuming that this pattern has a large effect on the beam halo, it would be an optimal candidate for a feasible HEL excitation pattern.

The resonant excitation in general features a rich nonlinear dynamics leading to a diversity in beam distribution and diffusion changes as can be seen for example from the very diverse effects of the different $k^{\mathrm{th}}$~turn pulsing patterns observed in simulations and experiments. The origin of this diversity is due to the fact that only certain frequencies and thus resonances are excited. Typical for the resonant excitation is also the fast adjustment of the beam distribution to a new equilibrium state which could be for the first time measured in the experiments and predicted in the simulations presented in this paper. The random excitation on the other hand excites all beam frequencies equally, leading to the highest increase in halo removal rates for the HEL \cite{hel_halo_hllhc_fitterer} but also the strongest effects on the core, mainly in terms of strong emittance growth as predicted in simulations, experiments and also a multitude of earlier publications on the subject of random noise.
